Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2023-02-23 Manuscripts/deNovoRates.tex
modified:   Manuscripts/Figures/pGainLossFix.svg 	new file:   Manuscripts/Figures/pGainLossFix2.svg 	modified:   Manuscripts/deNovoRates.tex 	modified:   Modeling/DeNovoProb.jl

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec, xcolor}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

@@

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}). Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible). We found that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present. This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence. This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}). Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed. Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively. We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/pGainLossFix.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

@@

\cmnt{Reproduction and selection determine the rate at which new genes spread in a population. To understand the combined effect of mutations, reproduction and selection we calculated two fixation probabilities. First is the probability that one individual gains a gene ($P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via population dynamics ($P_\textit{fix}$). The second probability describes the event that all individuals in a population have a gene, that is lost in one individual ($P_\textit{gene-loss}$), and eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$). We calculated the log-log ratio of the two probabilities under three scenarios, wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$). We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously. We found that under the neutral scenario, gene loss with extinction is still more likely than gene gain with fixation. However, the minimum length of ORF that causes two independent gene extinction as likely as a single gene fixation event, is more than 150 codons. This minimum number of codons increases with decreasing effective population size ($N_e$). When the gene is beneficial then gene extinctions are less frequent. This is true even for a marginally beneficial gene, such that in populations larger than 6400 individuals extinction of genes of any length is less likely than their fixation. Conversely, when the gene is mildly deleterious, its independent loss in more than two populations is more likely than its fixation in one population.}

@@

\cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first). In other words, the RNA-first hypothesis is more feasible. Likewise, a negative value of the ratio would suggest that the ORF-first hypothesis is more feasible. We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first (Fig SXXX). More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length. This is true for loci with three different values of GC content (34\%, 42\%, 50\%). However, our complementary analysis using DNA oligomer frequencies from \textit{D. melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length. Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.}

@@

To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$). We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}). At low GC-content (0.3 -- 0.34) ORF loss is less likely than RNA-loss, for short ORFs (30 codons; \hyperref[lossprob]{\autoref{lossprob}B}).

@@

We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths. Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.

Thus, widespread transcription loss can suggest a negative selection on protein expression.

@@

In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss. We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of proteins encoded by proto-genes. To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform. Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids. On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}{\color{blue}A}). This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid. However, GC-content also plays a role. For example, leucine, arginine and serine, all have six codons each. However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 0.42 (\autoref{aafreq}{\color{blue}A}). Overall, our analysis is roughly in agreement with many previous studies \citep{aasubOhta,aafreq3,aafreq4}. However, notable differences exist between our analysis and that of the previous studies because the latter primarily focused on characterized proteins and not random proto-gene derived protein sequences.

@@

We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).

@@

The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).

@@

As we mentioned in the previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations. We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}). The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836. That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other. For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution. This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection. We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}). Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that its median value is 0.38. Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.

@@

We found that $\sim$37.5\% of amino acids in a protein sequence are expected to be hydrophobic. This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}. We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}. Next, we asked if a protein sequence in general tends to be hydrophobic. Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).

@@

We found that the expected hydropathy is higher than zero ($0.39$), which suggests that random protein sequences are not on an average, hydrophobic.

@@

Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is 1.114, which suggests that gain of a hydrophobic amino acid is slightly more probable than its loss. We performed a complementary analysis where we calculated the average change in hydrophobicity due to any random mutation ($\bar{\Delta\alpha}$), defined by:

\begin{equation}
\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y}}
\end{equation}

We found the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity. We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein. To focus exclusively on non-hydrophobic to hydrophobic amino acid substitutions, we excluded amino acid pairs where both amino acids are either hydrophobic or are non-hydrophobic. That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity. With this focused analysis, we find that the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).

Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect is very small.

@@

We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge. In a hypothetical case, a proto-gene may be present in a species (A) but is absent in two sister taxa (B and C). Two evolutionary scenarios could explain this observation. In the first scenario, the proto-gene simply emerged for the first time in species A. The second scenario posits that the proto-gene was present in the common ancestor of the three species but was lost in both species B and species C. Our models suggest that the second scenario is more likely. Future phylogenetic studies on proto-genes should consider this finding for making inferences on the dynamics of gene gain and loss.

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. However, if both transcription and ORF are absent in a genomic locus, then \textit{de novo} emergence is more likely to occur transcription-first. Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene. For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory. On the other hand, the selection would have no effect on an ORF that is not transcribed. This is based on the assumption that the ORF sequence has no other function (for example, regulation), or is not under indirect selection (for example, if it overlaps with an existing gene). Thus positive selection will intensify the preference for RNA-first trajectory.

@@

Like all theoretical models, our models are not without limitations. First, we assume that nucleotides are uniformly distributed according to GC-content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}. The accuracy of the models' prediction will depend on the accuracy of parameters. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Fourth, we did not model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes. For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}. We instead focused on minimal requirements for transcription and translation. However, our modeling framework allows inclusion of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}). Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations. Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

\usepackage[usenames,dvipsnames]{xcolor}

@@

If a gene is found in one population but not populations, then the most parsimonious assumption is that this gene emerged only in one specific population. We asked what is the chance that the gene was present in an ancestral population but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}). Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible). We found that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present. This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence. This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}). Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed. Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively. We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:

@@

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/pGainLossFix2.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population. To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals. First is the probability that any any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$). The second probability describes the event that all individuals in a population have two copies of a gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$). We calculated the log-log ratio of the two probabilities under three scenarios, wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$). We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously. Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele. Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF. For example, a gene of any length is less likely to go extinct than it is fixed, in populations with at least 6400 individuals. Furthermore, a beneficial gene can be independently lost in more than two small populations with 400 individuals, only when it has more than 133 codons. When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population.}

@@

\cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first). In other words, the RNA-first trajectory is more feasible. Likewise, a negative value of the ratio would suggest that the ORF-first trajectory is more feasible. We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first (Fig SXXX). More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length. This is true for loci with three different values of GC content (34\%, 42\%, 50\%). However, our complementary analysis using DNA oligomer frequencies from \textit{D. melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length. Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.}

@@

\cmnt{To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$). We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}). If transcription is strictly dependent on a core promoter (TATA-box or Inr), small genes (30 codons) with low GC content ($>$34\%) can be preferentially lost by loss of transcription (Figure SXXX). When the organismal fitness is affected only by the protein product of a proto-gene, then both RNA loss and ORF loss would affect the fitness equally. Thus the likelihood of gene extinction through ORF loss or RNA loss depends solely on the probability of these events.

Our analysis of gene loss mechanism considers the loss of one feature (for example transcription) without requiring the other feature to remain intact (for example the ORF). With more stringent analysis where we consider the loss of only one feature but not the other, we find a the that exclusive ORF loss is more likely than exclusive RNA loss (Supplementary section XXX). The difference between the likelihood of the two gene loss mechanisms is more remarkable when we consider exclusive loss of one feature.}

Overall, our analysis suggests widespread transcription loss may be indicative of a negative selection on protein expression.

{\color{PineGreen} \textbf{MOVE TO SUPPLEMENTARY DATA}

@@

We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths. Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.
}

@@

In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss. We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of proteins encoded by proto-genes. To this end, we first asked if the expected frequency of different amino acids is uniform. For a GC-content of 42\% and with DNA trimer frequencies from \textit{D. melanogaster} intergenic genome, we found that it is not uniform. Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids. On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}{\color{blue}A}). This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid. However, nucleotide composition also determines the frequency of an amino acid. For example, leucine, arginine and serine, all have six codons each. However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 42\% (\autoref{aafreq}{\color{blue}A}). The same is true for the frequencies of these amino acids estimated from the DNA trimer distribution. Overall, our analysis is roughly in agreement with many previous studies \citep{aasubOhta,aafreq3,aafreq4}. However, notable differences exist between our analysis and that of the previous studies because the latter primarily focused on characterized proteins and not random proto-gene derived protein sequences.

@@

We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (GC = 42\%, Pearson $\rho = 0.955, P<10^{-10}$; \textit{D. melanogaster} trimer frequencies, Pearson $\rho = 0.841, P<10^{-10}$).

@@

The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (GC = 42\%, Pearson $\rho = -0.255, P = 0.2818$; \textit{D. melanogaster} trimer frequencies, Pearson $\rho = -0.356, P=0.1232$).

@@

As we mentioned in the previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations. We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}). The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836. That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other. For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution. This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection. \cmnt{We note that that actual frequencies of specific substitutions depend on the initial distribution of different codons in the ORF sequence.}

@@

We found that $\sim$40\% of amino acids in a protein sequence are expected to be hydrophobic (with a GC-content of 42\% as well as with \textit{D. melanogaster} trimer distribution). This finding is broadly in agreement with the previous study \citep{hydrophobicRatchet}. We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}. Next, we asked if a protein sequence in general tends to be hydrophobic. Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).

@@

We found that the expected hydropathy is higher than zero (0.38 with GC\% = 42; 0.36 with \textit{D. melanogaster} trimer frequencies), which suggests that random protein sequences are not on an average, hydrophobic.

@@

Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is slightly more than 1 (1.114 with GC\% = 42; 1.015 with \textit{D. melanogaster} trimer frequencies). This suggests that gain of a hydrophobic amino acid is slightly more probable than its loss. We performed a complementary analysis where we calculated the average change in hydrophobicity due to any random mutation ($\bar{\Delta\alpha}$), defined by:

\begin{equation}
\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} \mu_{x\to y}}
\end{equation}

We found the average change in hydrophobicity to be slightly less than zero ($-0.0798$), which suggests that on an average mutations cause a very small increase in hydrophobicity. We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein. To focus exclusively on non-hydrophobic to hydrophobic amino acid substitutions, we excluded amino acid pairs where both amino acids are either hydrophobic or are non-hydrophobic. That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity. With this focused analysis, we find that the average change in hydrophobicity is comparable ($-0.0793$). For both the analyses, we note that the actual change in hydrophobicity will depend on the composition of the ancestral protein sequence.

Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect may be very small.

@@

We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge. In a hypothetical case, a proto-gene may be present in a population (A) but is absent in two other geographically isolated populations (B and C). Two evolutionary scenarios could explain this observation. In the first scenario, the proto-gene simply emerged for the first time in population A. The second scenario posits that the proto-gene was present in the ancestral population but was lost in both population B and population C. Our models suggest that the second scenario is more likely than is assumed. \cmnt{This inference can be further extended to the level of closely related species where the divergence times are relatively shorter than the timescale of multiple gain and loss events within a lineage}. Future phylogenetic studies on proto-genes should consider this finding for inferring the dynamics of gene gain and loss.

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first). Long non-coding RNAs may be a potential source of proto-genes but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

CHECK THIS XXX: Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene. For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory. On the other hand, the selection would have no effect on an ORF that is not transcribed. This is based on the assumption that the ORF sequence has no other function (for example, regulation), or is not under indirect selection (for example, if it overlaps with an existing gene). Thus positive selection will intensify the preference for RNA-first trajectory.

@@

XXX: Like all theoretical models, our models are not without limitations. First, we assume that nucleotides are uniformly distributed according to GC-content. This is not true as nucleotide composition can vary significantly across the genome. Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}. However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci. Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}. The accuracy of the models' prediction will depend on the accuracy of parameters. However, mutation rate itself only determines the values of different probabilities. Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate. More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other. Thus our inferences based on our chosen parameters will not universally apply to all species. Third, we assume that all mutations are independent. This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}. Fourth, we did not model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes. For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}. We instead focused on minimal requirements for transcription and translation. However, our modeling framework allows inclusion of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}). Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation. Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations. Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.
